### 2016-09-25 oldmort.Rnw
add education and wages

%\usepackage[tablesonly]{endfloat}
\usepackage{endfloat}

@@

\section{Introduction}

In this study we take a long-term perspective on the development of social
inequalities in mortality in two Swedish region 1901--2013, in this case
with focus on the adult and elderly population. We investigate the development of
social inequality in mortality during the mortality transition and the
possible changes in the impact of social class in this process, in
particular the hypothesis that modern pattern of social differences in mortality appeared
during the first half of the 20th century.

Several studies have shown how important access to economic and other
resources are for health, but it is not clear how this has developed in
history. \citet{ant67} and \citet{smi83} suggest that social inequality in
health has passed through different phases in history. According to them,
social health differences were
comparatively small during the pre-transitional phase when space was a
strong determinant for the spread of disease. Differences then increased
during the transitional phase when mortality declined and wealthy groups
used their resources to gain better health. Finally, health differences
decreased again in modern low-mortality settings when instead
health-related behaviour became important, resulting in marginal
differences. \citet{omran82} state, in the third proposition in his theory
of the epidemiological transition, that even if the class differential in
mortality was maintained during the transition, the decline
set in earlier and was faster among privileged groups.

When studies, stimulated by the widely debated Black report \citep{whi98},
began to be published on the late 20th century development, any assumptions
of declining and even disappearing differences could be refuted. Even in
modern wealthy societies social health inequalities are substantial. In
fact, it even seems as if (at least relative) differences are increasing
\citep{kuetal04,frlu07}, contrary to what Antonovsky assumed. Research has
revealed a complex pattern of social class, material conditions,
psycho-social effects, and health-related behaviour. Many contemporary
studies implicitly or explicitly assume a "historical constant" of social
health inequality that often is considered to have been even greater in
earlier times \citep{mar04, liph95}.

It is however time to closely scrutinise the assumption about health and mortality inequalities in the past.
Our knowledge about social health inequalities in history, is
still based on scattered observations and usually not based on thorough
analyses. The social aspects of health began to be observed a long time ago. The 19th century French physician and pioneer in epidemiology Villerm\'e found large differences in mortality between city districts of varying social composition in early 19th century Paris, and
Engels gave evidence of appalling living conditions among the impoverished
in the rapidly expanding industrial environments \citep{ack48}. Despite
their merits as sources for 19th century society and their path-breaking
ideas, these 19th century observations are problematic in many ways. They
lacked adequate sources (no individual data) and methods for analysis. Another issue concerns
the degree to which we can generalise these observations to other
historical contexts.

During recent years, however, increased attention has been paid to the historical
perspective on social health inequalities. The results from different
studies have diverged, some reporting substantial differences but many
finding surprisingly small differences during the first stages of the
transition \citep{bevp11}. Nevertheless, it is certainly not obvious
that historical societies were characterised by the assumed positive
association between socio-economic position and survival. Several traits in
the pre- and early transitional society may explain this circumstance. The
effect of social class is dependent both on patterns of exposure and
resistance against diseases in different populations and sub-populations. In their overview on this topic \citet{bevp11} suggest that consistent hierarchical social health inequalities did not developed until the
first half of the 20th century.

Concerning Sweden, \citet{edli11} found no indication of better health
among wealthy adults living in the Sundsvall region during the
pre-industrial period 1803--1859. This pattern remained for men during the
industrial period 1860--1900, while clearer social health differences
appeared among women. A similar pattern was found for those in old age
\citep{edbr12}. This illustrates the importance of considering gender as
well as age patterns. Results from a study of other parts of northern Sweden,
the Skellefteå  and Umeå regions, 1851--1950 was in accordance with the results
from the Sundsvall region. Higher social position and more resources did not lead
to any health advantages for men, while this was the case for women \citep{segb14}. This is further illustrated with what is known about mortality in younger age groups in the same areas. \citet{edv92, edv04} found weak or
varying social differences in infant mortality in early 19th century northern Sweden, while industrialisation caused large social differences in mortality among children 1--14 years. In their study on socio-economic
mortality differences in a couple of Scanian parishes 1815--1968,
\citet{bedr11} found no indications of any social gradient during most of
the period, suggesting that the increase is a recent phenomenon.

To summarise the present state of knowledge about the development of social
health inequalities (in Sweden but also in other European countries), no
clear social hierarchy has been found before 1900 but substantial
differences towards the late 20th and early 21st century. It appears as if
a more consequent social health divide developed in the period
in-between. This period has not been thoroughly studied on this issue,
mainly because of lack of data to analyse. However, the extension of data at the Demographic Data Base, Umeå University (DDB) makes such studies possible. We are now able to analyse social
differences in mortality with micro-data on a sufficiently large population and with substantial social diversity from the 20th century for the Skellefteå and Umeå regions \citep{wee16}.

\section{Aims and questions}

The aim of the paper is to investigate the development of social
differences using HISCLASS in mortality for the adult and elderly population during the mortality
transition and its relation to the development of income inequality. The analyses
will be performed on data from the Skellefteå and Umeå regions
1901--2013. The questions are:

\begin{enumerate}
\item Has inequality in mortality between social classes increased among the adult and elderly
  population, primarily benefiting the highest social positions?

%%\item Is the mortality transition initiated by a certain social class?

\item Are there any gender differences in the effect of social position?

\item Is social position equally important among the retired population as for those in working age?
\end{enumerate}

\section{Theoretical considerations}

The WHO Commission on Social Determinants of Health framework for the
social determinants of health discusses how socio-economic position impacts
on the health of individuals, placing them within structures that determine
their opportunities and how these influences works
through intermediate determinants related to exposure and vulnerability
\citep[p. 48]{who10}. People are situated within socio-economic and political
contexts, as well as value systems in the society. The socio-economic
position imply different access to resources, power and prestige, all
aspects of crucial importance. The framework takes a holistic approach to
the social determinants, including not only the effect of pure economic and
material resources. The importance of understanding people in their full
context has been emphasised by for example \citet{kri05}. What she calls embodiment refers to the recognition that human beings are
simultaneously social beings and biological organisms.

Social inequality can mean different things, and studies of social
differences 
in health and mortality often differ in what sort of categorisation is used.
This is important to have in mind when comparing results, because the chosen
social variable may measure different aspects of inequality. This relate to an
old subject within social studies, on class and status variables, based on
forerunners 
such as Marx and Weber. In many modern studies, data on income, wealth
and/or education are frequently used. The underlying assumption is that more access
to these resources leads to advantages in life, for example when it comes to survival.
SOCPO \citep{vdpm05} is a system based on different access to what they call social power.
\citet{goldk10} criticise how many economists and epidemiologists operationalize
inequality in their studies, for example by not distinguishing between class
and status aspects. Class refers to relational aspects in society, while many other
variables focus on the distribution of material and/or immaterial resources.
He advocates studies of different aspects, which together will complement each other.

HISCLASS, the classification system used here, is a "\ldots HISCO-based
historical international social class scheme" \citep[p.\ 26]{lema11}. The
different classes in HISCLASS represent distinct categories, based on
whether the work is manual or non--manual, skill level, supervision and
sector and it does not represent a purely hierarchical system, 
but implicitly it reflects large differences in what is in the forefront of
other categorisations, 
as for example access to economic and other resources, status, social power et cetera.
We can thus assume that classes have different opportunities when it comes to matters of health and mortality.

A reasonable assumption is that levels of economic inequality in a society
influences social patterns of health and mortality. Social classes with
more economic resources have better possibilities to stay healthy. We
expect that people from classes with less resources are seriously
disadvantaged in a society with large income inequality. Their poverty
would expose them to larger health risks. The inequality effect could
however influence health in other ways than at the individual economic
level. \citet{wipi09} suggest that income inequality have a contextual
effect as well, meaning that living in an unequal society increases health
risks above what is explained by the individual situation.

How socio-economic position influences lives of people, their health and
their mortality needs to be problematized. When it comes to differential
access to resources, these could be of different character, and not only
material ones. \citet{bou84} distinguishes between economic, 
social and cultural capital. Economic capital allows for much more freedom
to organise lives and to maximise life opportunities. Social capital (here
referring to its individual variant and not as a description on society
level) include different kinds of networks, for example family, friends,
colleagues and other persons acting as valuable contacts and resources. A
rich social network can be crucial when it comes to life opportunities. The
impact of cultural capital on health may seem less obvious. Cultural
capital refers both to the personal skills but also to prestige and
status. This can be demonstrated in taste and choice of life style, which
bring people together with others of similar background and
interests. Achieving a higher cultural capital involve several requirements
on way of living, something that can be relevant for health. Cultural
capital in itself can also bring beneficial psycho-social effects on
individuals, which could be an advantage for health \citep{mar04, wipi09}.

In the historical part of this paper, individuals are grouped
according to their family households, that is, children and married women are
grouped according to the occupation of the head of household, usually the
father or husband. A practical reason for this is that occupation is usually
only provided for the head of household, but there are good research
motives for this as well. We consider the whole family to belong to the
same socio-economic position. But this does not mean that we necessarily
believe that the effects are similar for both sexes. As several researchers
have pointed out, demography has often ignored power relations and
different roles within families \citep{watkins93, mack95, jans07}. By
investigating possible differences in mortality patterns 
according to sex, we may identify gendered patterns within the family, resulting in different effects on family members. Different patterns between the sexes
may be caused by discrimination of the female part in the household. We
know that the society was strongly segregated according to gender and that
women were discriminated and had less space for action. Another possible
cause for different patterns is the gendered expectations on behaviour that
has characterised the relation between sexes in societies throughout
history.
Unfortunately, the same approach cannot be applied on the data from 1960 onward, due to difficulties identifying couples in the modern data. This is further developed in the section on the construction of data for analysis.

\section{Historical background}

The historical context of the present study is a period of major changes in
Sweden and in the studied regions. Swedish society underwent an enormous
development, changing from a rural and sparsely populated society based on
agriculture and with low economic standard to a modern, industrialised and wealthy country. Material conditions improved and welfare provisions for people increased, mainly funded by public finances. Several traits of what was to become the Swedish welfare state had early
predecessors, but became even stronger from the 1930s. These reforms
fundamentally changed living conditions for all citizens but most of all for the
most vulnerable, for example the elderly. Still, levels of income inequality differed during the studied period. The early 20th century was characterised by large income
and wealth differences. This started to decline from the 1930s, but it was
still high into the 1950s. It was only in the
following decades that income inequality became very small in Sweden,
making Sweden one of the most (or even the most) equal societies in the
world around 1980 when inequality levels seem to have been at the lowest. During recent decades, inequality has however increased again \citep{pik14, rowa08, rowa09}.

At the middle of the 19th century, Sweden belonged to the part of the world
with the best survival, being number one in the 1840s, and has continued to belong to the countries with lowest mortality.  Life expectancy was 56 years 1901--10, remaining life
expectancy at age 50 was 24 years. This increased to 72 versus 27 in 1951-59 and 82 versus 33 in 2013. Younger age groups contributed most to the initial and total decline, while improved survival in old age groups adds most to more years during recent decades. The First World War marks an
important breaking point towards declining mortality among young adults, while the period after the Second World War led to large improvements for older adults. \citep{holu76}. If spatial differences determined mortality during the pre-transitional, regional variation and urban
rural differences in mortality levels have diminished, although not disappearing completely.

We do not argue that the Skellefteå and Umeå regions represent the whole of
Sweden, but the availability of data allows us to study the transition
in greater detail here making it a laboratory for studying Swedish demography during a period of fundamental changes. Both regions are part of the county of Västerbotten in the north of Sweden,
along the coast of the Gulf of Bothnia. The Skellefteå region in Poplink consists of a selection of
parishes surrounding the town of Skellefteå, founded in 1845 but with a
very small population in the 19th century. The data from the period after 1960 cover the Skellefteå and Malå municipalities, the same area as for the earlier period but with the addition of two more parishes. The majority of the population in the region lived in rural villages and hamlets, getting their
livelihood from agricultural production. The
region was vulnerable for harvest failures, one living long in memory was
the famine of 1867--1868 \citep{segb14}. Towards the end of the
century, the region became connected to the Swedish railway system, thus
improving the communications and contacts with other parts of the
country. During the 20th century, industrialisation took place, partly by
the opening of mines in the inlands. This also led to a population increase both in the town and in the rural parts,
and a much more diversified economy.

<<skpop>>=
library(skum)
ske1850 <- with(obs[obs$region == "ske", ],
                sum(birthdate + enter < 1850 &
                    birthdate + exit > 1850))
ske1900 <- with(obs[obs$region == "ske", ],
                sum(birthdate + enter < 1900 &
                    birthdate + exit > 1900))
ske1950 <- with(obs[obs$region == "ske", ],
                sum(birthdate + enter < 1950 &
                    birthdate + exit > 1950))

@

Mortality was fairly low in comparison
with other parts of the country and the fertility transition was late. The
Skellefteå population size as defined in our data-sets was \Sexpr{ske1850} on January 1, 1850,
\Sexpr{ske1900} on January 1,
1900, \Sexpr{ske1950} on January 1, 1950 and 76723 at the end of the century.

\begin{figure}[ht!]
\includegraphics[height=19cm]{myfigs/sweden}
\caption{Skellefteå and Umeå in Sweden.}
\end{figure}

%%\section{Map. Skellefteå and Umeå regions}

<<umpop>>=
ume1900 <- with(obs[obs$region == "ume", ],
                sum(birthdate + enter < 1900 &
                    birthdate + exit > 1900))
ume1950 <- with(obs[obs$region == "ume", ],
                sum(birthdate + enter < 1950 &
                    birthdate + exit > 1950))
@

The Umeå region in Poplink (until 1950s) consists of Umeå urban and rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included. This region had a somewhat different character from that of Skellefteå, even though agriculture was
dominating the economy for a long time. The town of Umeå had a small population but
substantially larger than Skellefteå town during almost the whole studied
period. It was the administrative centre in the county of Västerbotten and
schools and military regiments was placed here. Officials, teachers and the
military were much more common in Umeå. Agriculture dominated in the rural part, while  and there were some foundries, some industries for example in forestry and small-scale production. Umeå had a much more diversified
occupational structure than Skellefteå. The population size as defined in our data-sets was
\Sexpr{ume1900} on January 1, 1900, \Sexpr{ume1950} on January 1, 1950 and 103970 when the century ended.

\section{Data and variables}

The data for the present study comes from two large population databases at
the Demographic Data Base, Umeå University. The early period is covered by
the database Poplink, the digitisation of historical parish registers for
the Skellefteå and Umeå region \citep{wee16}. The population in the two
regions is large enough and socially diverse to enable studies of social
mortality inequalities. Poplink is based on linked parish records, allowing
us to reconstruct life biographies on people as long as they remained in
the region. The records are linked within but not between the regions. Data
from the regions are accessed for the period 1901--1950s.

The other large data-set comes from the Linnaeus database
\citep{malm10}. It is based on different linked population registers from
1960 to 2013 and is used within the ageing program at CEDAR, Umeå
University. The study period from 1960 to 2000 are constructed from censuses every fifth
year 1960--1990, with additional information on deaths from National Board of Health and Welfare. For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare. The construction of these data-sets for analysis is presented below.

Information on individuals are anonymized and the two databases are not
linked between each other. This stops us from following individuals present
in both databases throughout their lives, thus they are treated separately. This
also makes it impossible to add information on individuals in the Linnaeus
database from what we potentially could find in Poplink, for example family
background or occupations earlier in life.

In the data-set analysed here, all individuals 40 years and older and ever being resident in either
of the regions are included. The data file contains the variables on sex,
birth date, death date, first and last date of observation and type of
entrance/exit.

The identification of presence periods as well as the time-dependent variables require some additional comments.

\subsection{Presence periods}

Differences in available information, both when it comes to structure and variables used for attributing social class, from the two data-sets as well as
within the Linnaeus data-set makes necessary diverging approaches when it comes to
the identification of presence periods within our
regions. The Poplink data provides us with exact dates or at least year of
start and exit of the presence of all individuals in our data. This is however not the case
with the Linnaeus database. For the period 1960--2000 we use the presence
in the censuses 1960--1990 together with the yearly population registers
1986--2000 for defining the observation periods as well as their social
class (see below) in our analyses. We have however no exact information on
what happened in between the censuses, for example the exact date of exit
for persons that moved between two censuses. We have solved this by
considering everyone as resident in our regions during the complete
five-year period between censuses if the person is missing in this
census.This overestimates their presence in our regions, but should not
bias the results. We furthermore miss those that both moved in and moved out
or vice versa between censuses. Exit dates are exact if they refer to
death, and in these cases we consider them as constantly present from the
last census to the death data even if the death took place in another part
of Sweden.

For the last period, 2002--2013, we use the yearly population registers
(also known as LISA data from Statistics Sweden), together with information
on deaths from National Board of Health and Welfare. This data allows a
better control of the presence periods in the studied regions.

The presence periods, with start and end date, are reconstructed and
divided according to two time-dependent variables: Social position and
marital status. Thus, at each change in any of
these variables, a new record is created. The time-dependent variables require some additional comments.

\subsection{Social position}

In this paper we analyse mortality differences according to social class
according to the classification scheme HISCLASS. The only type of
socio-economic information available both in Linnaeus and Poplink is
occupations, which can be used to identify distinct social classes or
social status groups. Thus, analysing according to social class is apart
from a scientific choice, also a practical one. Using occupational data is
the only and obvious alternative if we are to make a longitudinal study of
socio-economic mortality differences with consistent classification schemes
(although not without some problems). There are also other problems
connected with using the same classification scheme covering a long time
period, but this will be approached in more detail in the discussion
section. Social position is treated as a time-dependent variable.

The availability to information on occupations varies over time and between sources. The Poplink data usually provides information on the occupation of head of households. The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering). The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1990 to 2001 where we lack information. In the censuses, we have solved this by extending the periods for which we consider the defined social class be valid. Thus, the social class in the census 1990 is considered to be kept until 2000.

The use of different data sets, forces us to apply different approaches towards defining the social position. In the Poplink data, social class is based on the occupational titles given in the parish records. These can be more or less precise. There are some problems with titles that are not precise enough to identify the size of the enterprise, or the division between skills, semi-skilled and unskilled workers. We consider however their validity to be sufficient for identification of social position. Another problem is that we in Poplink usually get the occupation only for the head of household, thus under-estimating the female working-force participation and the occupations of adult children residing with their parents \citep{vik10}. In this study and according to the research questions in the present paper, we have chosen to categorise wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position. This strategy makes the problem of under-recording less problematic.

Female labour is however much better covered in the Linnaeus database, both in the censuses and the population registers. Another difference is that it is much more difficult to define households in the same way as in Poplink from the Linnaeus database. Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became very common in Sweden during this period. It is also worth to notice that female labour force participation became the norm, and we could therefore argue that the occupation of the male as signifier of the social class became obsolete. We therefore let all included persons be signified by their individual occupation for this period. Still, we need to keep in mind that results for the different periods are not completely comparable.

Due to the construction of the time-dependent data, some records report unknown social position. This has been updated according to rules in order to optimise available information and minimise missing information. Those that still lack information on social class are not included in the analysis, something that have consequences for elderly living in the 1960s in particular, since many were already retired and thus without occupation when we first observe them in the 1960 census. The fact that Linnaeus and Poplink are separate databases stops us from updating this information from Poplink.

The classification scheme used here is HISCLASS \citep{lema11}, a class-based system. The scheme uses HISCO as a basis for assigning categories. HISCO \citet{lema02} (Historical International Standard Classifications of Occupations) is a classification system of historical occupations, representing a historical version of ISCO. HISCO groups occupations according to tasks and skills.
HISCLASS have 12 categories, but some categories become quite small in number even when analysing these comparatively large regions. This is particularly apparent when analysing the development in separate periods where for example farm workers become more and more rare during the studied century. We have therefore collapsed some categories, ending up with the following six:

\begin{enumerate}
\item \'Elite, HISCLASS 1 and 2, Higher managers and higher professionals.
\item Lower managers, HISCLASS 3.
\item Farmers and fishermen, HISCLASS 8.
\item Lower professionals, HISCLASS 4, 5 and 6, Lower professionals, lower clerical, foremen.
\item Medium skilled workers, HISCLASS 7.
\item Lower skilled workers, HISCLASS 9, 10, 11 and 12, Lower and unskilled workers (including farm workers)
%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
\end{enumerate}

In the last studied period, we have brought together the categories even more.

For the Poplink data, we have used the script recoding HISCO into HISCLASS. The procedure for the occupations in the Linnaeus database is a little more complicated. The occupations in the censuses (NYK) must first be transferred to the system later used, SSYK (which is the Swedish version of ISCO88). Thereafter the SSYK, both those transferred from the censuses and those given in the population registers 2001-2013 can be recoded into HISCLASS. We have used a script provided by Jonas Helgertz, Lund University for this transformation.

\subsection{Marital status}

Marital status is treated as a time-dependent variable in the data file. The statuses used are unmarried, married and dissolution of marriage. The status unmarried is often not explicitly stated, which makes the status unknown in the database Poplink for people for people moving to our parishes.  If no explicit status is given in the sources and partner is missing, the status has been set to unmarried in our data file.
%'
%'
%' <<getdata, results='asis'>>=
%' library(xtable)
%' library(knitr)
%' load("data/adult.rda")
%' @
%'
%'
%' <<toonerec>>=
%' adult <- rc(adult)
%' ##mad <- adult
%' ##indx <- with(mad, tapply(id, id))
%' ##mad$Enter <- with(mad, tapply(enter, id, min))[indx]
%' ##mad$Exit <- with(mad, tapply(exit, id, max))[indx]
%' ##mad$Event <- with(mad, tapply(event, id, max))[indx]
%'
%' ##mad <- mad[mad$lopnr == 1, ]
%' ##mad$enter <- mad$Enter
%' ##mad$exit <- mad$Exit
%' ##mad$event <- mad$Event
%' ##mad$Enter <- mad$Exit <- mad$Event <- NULL
%'
%' ##mad <- mad[, c("id", "enter", "exit", "event", "birthdate",
%'   ##             "sex", "region",
%'     ##           "socpo20", "socpo30", "socpo40", "socpo50",
%'       ##         "civst20", "civst30", "civst40", "civst50",
%'         ##       "ortnmn")]
%'
%'
%' ## General:
%' source("R/addPer.R")
%' cut <- c(1851, 1876, 1901, 1921, 1940, 1951)
%' xc <- numeric(length(cut) - 1)
%' for (i in 1:(length(cut) - 1)){
%'   xc[i] <- (cut[i] + cut[i + 1]) / 2
%' }
%' ucut <- cut[-(1:2)]
%' uxc <- xc[-(1:2)]
%' mud <- addPer(adult, cuts = cut)
%' source("R/getInter.R")
%'
%' library(xtable)
%' ##xtable(summary(adult[, c("sex", "region",
%'   ##                       "civst40", "socpo40")]))
%' ##save(adult, file = "adult.rda")
%' ##save(mad, file = "mad.rda")
%' @
%'

\section{Models}

The proportional hazards model in a survival analysis context is used,
allowing for adjustment for civil status and rural/urban environment. Since
the central explanatory variable, {\sc Hisclass}, do not follow the
proportionality property, the analyses are stratified with respect to the
variable, and the main results are presented graphically.

The analyses are performed in the \R\ environment for statistical computing
and graphics \citep{cran}, especially using the package {\tt eha} \citep{gb12,eha}
The proportional hazards model allow us to scrutinise the differences and
the development in more detail. We divide the analysis into two age groups,
40--59 and 60--99 years of age. The model controls for period, marital
status and whether the individual resides in the urban or rural part. There
are no serious signs of deviation from proportionality assumption, except
regarding {\sc Hisclass}.

We analyse both regions together but perform separate analyses for men and
women and for the age groups 40--59 and 60--89. The model in the analyses
include social class, marital status and urban/rural residence. Each decade
during the 20th century gets separate analyses, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period. We present the results for the period 2001--2013
separately. We suggest a heterogeneity index in order to investigate a
possible homogenisation of mortality between social classes.

The important explanatory variable, {\sc Hisclass}, is included in the
proportional hazards models as a \emph{stratification} variable. The main
reason for this choice is that the effect of {\sc Hisclass} on survival is
non-proportional, that is, it varies with age. Thus, the main results are
presented \emph{graphically}. Note further that {\tt Hisclass} is a
\emph{time-varying} variable. This causes lack of information in the higher
ages around 1960--1980, because that kind of information is simply missing
to a large extent.

%The effect of *hisclass* on survival prospects is in focus, and the obvious
%tool to use is a proportional hazard (PH) model. The {\sc hisclass} level could
%either be analysed as a time dependent variable, or as a fixed variable at
%a certain point in time. In our case, {\sc hisclass} is defined at the age when
%we start to analyse mortality, age 40 or 50. At these ages, most have
%reached their top position and they are still work active. We thus analyse
%future survival according to the social position in mid-life. It has
%advantages when it comes to drawing causal conclusions of observed
%observations. Changes in social position may be due to circumstances that
%are related to mortality risks, for example changing occupation due to bad
%health. Another advantage of letting the {\sc hisclass} level be fixed is that
%time-dependent are more sensitive to the assumptions in a proportional
%hazard model. In our preliminary analyses of letting {\sc Hisclass} be
%time-dependent, we encounter serious deviations from the PH
%assumption. Marital status is however time-varying.

%As a first crude measure of survival, we use median remaining life at age 40. 
%The reason for showing the median and not the average remaining life is that the latter is impossible to estimate
%nonparametrically without strong assumptions. This depends of course on the
%kind of data we have: Heavily censored and truncated. Our approach can be
%seen as a hybrid between period and cohort analysis.The model is calculated
%for 20-year periods with the variables *hisclass*, marital status and urban
%versus rural residence in the model, and separately for the two regions and
%gender. The reason for presenting the regions separately is that there is
%no data covering the 19th century for the Umeå region. The results
%from the different analyses are presented in a serie of figures. We have
%also made statistical analyses of possible changes in the effect of
%*hisclass*, in order to identify possible appearance of a social
%gradient. These analyses are made separately for the age group 40-59 and
%age group 60 and above.

\subsection{POPLINK data, 1901--1950}

The POPLINK data cover 1901--1950.

\subsection{FOB data, 1961--2000}

The \emph{Fob} data is treated similarly as the POPLINK data. There is
unfortunately no links between the individuals in the two data frames, and
the result of that is a gap in the relevant information between the years
1950 and 1960. Furthermore, the first two decades in the \emph{Fob} data,
1960--1980, will have a lot of missing information on {\sc Hisclass}.

\subsection{LISA data, 2002--2013}

The {\sc Lisa} data is yearly, with {\tt hisclass} information starting in 2002. This information, however, do not allow us to contemplate seven (or six) levels, so we are using only four classes, grouping together 1+2, 3+4, 5, and 6.

@@

<<readtab>>=
load("../CreateData/data4/vbTab.rda")
@

@@

Models are fitted separately for each ten-year time period, from  1901 to
2000, with the exception of the period 1951--1960. We stratify on HISCLASS
and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
(Skellefteå town or outside). The cumulative
hazards at the end of the time period (60 if 40--59, 90 if 60--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

@@

\subsection{Ages 40--59}

%The distribution over time periods.
<<getske4060>>=
cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, women."
scap1w <-  "Cumulative hazards, ages 40--59, women."
cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, men."
scap1m <-  "Cumulative hazards, ages 40--59, men."
##
cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap12 <- "Total hazard by HISCLASS, ages 40--59."    
cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
scap12log <- "Total hazard by HISCLASS, ages 40--59, last years."

cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, women."
scap2w <-  "Cumulative hazards, ages 60--89, women."
cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, men."
scap2m <-  "Cumulative hazards, ages 60--89, men."
##
cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap22 <- "Total hazard by HISCLASS, ages 60--89."    
cap22log <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
scap22log <- "Total hazard by HISCLASS, ages 60--89, last years."

##
vb4059 <- vbTab[vbTab$age %in% seq(40, 55, by = 5), ]
vb4059$age <- factor(vb4059$age)
@

@@

<<hazcalw>>=
resw <- getPHglm(vb4059[vb4059$sex == "female", ],
            nper, labb)
resm <- getPHglm(vb4059[vb4059$sex == "male", ],
            nper, labb)
@

<<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
source("R/plotHaz.glm.R")
plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
@

@@

<<cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(60|40)",
                main = "Women")
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(60|40)",
                main = "Men")
par(oldpar)
@

%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--59 per individual by HISCLASS and decade, men.">>=
%%@

@@

<<cumslogw, fig.cap = cap12log, fig.scap = scap12log, fig.height = 10>>=
oldpar <- par(mfrow = c(2, 1))
wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(60|40)",
main = "Women")
mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(60|40)",
main = "Men")
par(oldpar)
@

The analysis for the last period is based on a separate analysis and is presented separately.

<<lissyoung>>=
load("../CreateData/data4/liss.rda")
liss4059 <- age.window(liss, c(40, 59))
@

<<liswom, cache = TRUE>>=
fitw1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, dist = "gompertz", data = liss4059[liss4059$sex == "female", ])
fitm1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, dist = "gompertz", data = liss4059[liss4059$sex == "male", ])
dcol <- c("blue", "darkgreen", "darkgreen", "red")
@

See Figure~\ref{fig:plotliss1} for the cumulative hazards.

>>>>

%\usepackage[tablesonly]{endfloat}
\usepackage[figuresonly]{endfloat}

@@

{\huge Här ska 'intro.tex' in !!!!!!!!!!!!!!!!!!!!!}

@@

<<readtab>>=
load("data/vbTab.rda")
@

@@

Models are fitted separately for each ten-year time period, from  1901 to
2000, with the exception of the period 1951--1960. We stratify on HISCLASS
and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
(Skellefteå town or outside). The cumulative
hazards at the end of the time period (60 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

@@

\subsection{Ages 40--64}

%The distribution over time periods.
<<getske4060>>=
cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women."
scap1w <-  "Cumulative hazards, ages 40--64, women."
cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men."
scap1m <-  "Cumulative hazards, ages 40--64, men."
##
cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap12 <- "Total hazard by HISCLASS, ages 40--64."    
cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
scap12log <- "Total hazard by HISCLASS, ages 40--64, last years."

cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women."
scap2w <-  "Cumulative hazards, ages 65--89, women."
cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men."
scap2m <-  "Cumulative hazards, ages 65--89, men."
##
cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap22 <- "Total hazard by HISCLASS, ages 65--89."    
cap22log <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
scap22log <- "Total hazard by HISCLASS, ages 65--89, last years."

##
vb4064 <- vbTab[as.numeric(as.character(vbTab$age)) < 64.5, ]
vb4064$age <- factor(vb4064$age)
@

@@

<<hazcalw>>=
source("R/getPHglm2.R")
resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
            nper, labb)
resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
            nper, labb)
@

<<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
source("R/plotHaz.glm.R")

plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
@

@@

<<cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
                main = "Women", exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
                main = "Men", exclude = 5:6)
par(oldpar)
@

%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
%%@

@@

<<cumslogw, fig.cap = cap12log, fig.scap = scap12log, fig.height = 10>>=
oldpar <- par(mfrow = c(2, 1))
wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(65|40)",
main = "Women")
mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(65|40)",
main = "Men")
par(oldpar)
@

%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH

\subsubsection*{The effect of income and education, 1990--2005}

The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables.

<<lissyoung>>=
library(eha)
library(xtable)
load("data/loTab.rda")
loTab$period <- factor(loTab$period)
liss4064 <- loTab[loTab$age < 64.5, ]
@
